---
title: "Final_project-  trashwheel"
author: "Ferencz Boróka"
date: "2024-12-15"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, echo=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading libraries
```{r }

library(tidyverse)
library(skimr)
library(lmtest)
library(car)
```

## Reading the file
```{r }


#reading the file with tidytuesdayR 
tuesdata <- tidytuesdayR::tt_load(2024, week = 10)

trashwheel <- tuesdata$trashwheel

```

## Inspecting the dataframe

```{r }

str(trashwheel)   # Check the structure
dim(trashwheel)   # Dimensions 
head(trashwheel)  # Preview first few rows
colnames(trashwheel) # List column names
sapply(trashwheel, class) # Check data types
table(trashwheel$Month) # Checking the month variable values

```

# Cleaning the dataframe

```{r }

# Count missing values per column
colSums(is.na(trashwheel)) 

# Impute or remove missing values (example: imputation for numeric columns)
trashwheel <- trashwheel %>%
  mutate(across(where(is.numeric), ~ ifelse(is.na(.), mean(., na.rm = TRUE), .)))

# Remove duplicates
trashwheel <- trashwheel %>%
  distinct()

# Convert Year to a factor
trashwheel$Year <- as.factor(trashwheel$Year)
unique(trashwheel$Year)

# Convert the Month column to lowercase so I'll only have 12 types
trashwheel$Month <- tolower(trashwheel$Month)

# Check the unique values to verify there are now only 12 months
unique(trashwheel$Month)

#making a YearMonth variable
trashwheel$YearMonth <- paste(trashwheel$Year, trashwheel$Month, sep = "-")

#Checking unique values
unique(trashwheel$YearMonth)
unique(trashwheel$HomesPowered)
unique(trashwheel$Name)
```


# Exploratory Data Analysis

```{r }
# Summary statistics
skim(trashwheel)


#making a new dataframe with a YearMOnth grouping and  averaging volume and wheight
trashwheel_volume_weight <- trashwheel %>%
  group_by(YearMonth) %>%
  summarise(
    mean_volume = mean(Volume, na.rm = TRUE),
    mean_weight = mean(Weight, na.rm = TRUE),
    .groups = "drop"
  )


# Plot Mean Weight over Time
ggplot(trashwheel_volume_weight, aes(x = YearMonth, y = mean_weight, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(title = "Mean Weight Over Time", x = "Year-Month", y = "Mean Weight") +
  theme_minimal() +
  scale_x_discrete(breaks = trashwheel_volume_weight$YearMonth[seq(1, nrow(trashwheel_volume_weight), by = 12)]) # Show every 12th value

#plot Mean Volume over Time
ggplot(trashwheel_volume_weight, aes(x = YearMonth, y = mean_volume, group = 1)) +
  geom_line(color = "red") +
  geom_point(color = "red") +
  labs(title = "Mean Volume Over Time", x = "Year-Month", y = "Mean Volume") +
  theme_minimal() +
  scale_x_discrete(breaks = trashwheel_volume_weight$YearMonth[seq(1, nrow(trashwheel_volume_weight), by = 12)]) # Show every 12th value

#making another dataframe with Yearmonth and Name grouping and homespowered averaged
trashwheel_homespowered <- trashwheel %>%
  group_by(YearMonth, Name) %>%
  summarise(
    mean_homespowered = mean(as.numeric(HomesPowered), na.rm = TRUE), # Convert HomePowered to numeric if needed
    .groups = "drop"
  )

#Plot: homespowered over time, with distinct trashwheels
ggplot(trashwheel_homespowered, aes(x = YearMonth, y = mean_homespowered, group = 1)) +
  geom_line(color = "blue") +
  geom_point(color = "blue") +
  labs(
    title = "Mean Homes Powered Over Time by Trashwheel",
    x = "Year-Month",
    y = "Mean Homes Powered"
  ) +
  theme_minimal() +
  facet_wrap(~ Name)

# Group by YearMonth and calculate means for specified trash types
trashwheel_summary <- trashwheel %>%
  group_by(YearMonth) %>%
  summarise(
    PlasticBottles = mean(PlasticBottles, na.rm = TRUE),
    Polystyrene = mean(Polystyrene, na.rm = TRUE),
    CigaretteButts = mean(CigaretteButts, na.rm = TRUE),
    GlassBottles = mean(GlassBottles, na.rm = TRUE),
    PlasticBags = mean(PlasticBags, na.rm = TRUE),
    Wrappers = mean(Wrappers, na.rm = TRUE),
    SportsBalls = mean(SportsBalls, na.rm = TRUE),
    .groups = "drop"
  )

# Reshape the data to long format 
trashwheel_long <- trashwheel_summary %>%
  pivot_longer(
    cols = c(PlasticBottles, Polystyrene, CigaretteButts, GlassBottles, PlasticBags, Wrappers, SportsBalls),
    names_to = "TrashType",
    values_to = "MeanAmount"
  )

# Plot: trends of trash types over time
ggplot(trashwheel_long, aes(x = YearMonth, y = MeanAmount, color = TrashType, group = TrashType)) +
  geom_line(size = 1.2) +
  scale_y_log10(labels = scales::comma_format()) + # Logarithmic scale for y-axis
  scale_x_discrete(breaks = trashwheel_summary$YearMonth[seq(1, nrow(trashwheel_summary), by = 12)]) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Trends of Trash Types Over Time (Logarithmic Scale)",
    x = "Year-Month",
    y = "Mean Amount (Log Scale)",
    color = "Trash Type"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels
  )


```

# Statistical analysis

## Building a simple model
```{r }

trash_model <- lm(Weight ~ PlasticBottles + Polystyrene + CigaretteButts + 
                  GlassBottles + PlasticBags + Wrappers + SportsBalls, 
                  data = trashwheel)

# View the model summary
summary(trash_model)

```


#Checking all assumptions for the simple model
## Checking Influential Points

```{r influential_points_trash, echo=TRUE}

# Check for influential points using Cook's distance
cooks_d_trash <- cooks.distance(trash_model)

# Visualize Cook's Distance
cooks_data_trash <- data.frame(Observation = 1:length(cooks_d_trash), Cooks_Distance = cooks_d_trash)
ggplot(cooks_data_trash, aes(x = Observation, y = Cooks_Distance)) +
  geom_point() +
  geom_hline(yintercept = 4 / length(cooks_d_trash), color = "red", linetype = "dashed") +
  labs(title = "Cook's Distance for Influential Observations",
       x = "Observation",
       y = "Cook's Distance") +
  theme_minimal()

# Identify observations with Cook's distance greater than 4/n
influentials_trash <- which(cooks_d_trash > (4 / nrow(trashwheel)))
influentials_trash
```


###Checking normality of residuals
```{r }
# Calculate residuals
residuals_trash <- residuals(trash_model)

# Plot histogram of residuals
ggplot(data.frame(residuals_trash = residuals_trash), aes(x = residuals_trash)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "blue", alpha = 0.7) +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Frequency") +
  theme_minimal()

# Q-Q Plot for residuals
qqnorm(residuals_trash)
qqline(residuals_trash, col = "red", lwd = 2)
```

### Checking linearity
```{r }
# Get the fitted values
fitted_values_trash <- fitted(trash_model)

# Create a data frame for residuals vs fitted values
residuals_vs_fitted_df_trash <- data.frame(fitted_values_trash, residuals_trash = residuals(trash_model))

# Residuals vs Fitted plot
ggplot(residuals_vs_fitted_df_trash, aes(x = fitted_values_trash, y = residuals_trash)) +
  geom_point(color = "skyblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()
```

### Checking Homoscedasticity
```{r }


# Scale-Location Plot for Visual Check
plot(trash_model, which = 3)

# Breusch-Pagan Test for Homoscedasticity
bptest_trash <- bptest(trash_model)
bptest_trash


```


### Checking multicolinearity
```{r }


# Calculate and check the VIF for the model
vif_values_trash <- vif(trash_model)

# Print VIF values
print(vif_values_trash)


```


## Modifying the simple model
```{r }
trashwheel$log_Weight <- log(trashwheel$Weight)

# building the logarthimical mdoel
trash_model_log <- lm(log_Weight ~ PlasticBottles + Polystyrene + CigaretteButts + GlassBottles + PlasticBags + Wrappers + SportsBalls, data = trashwheel)

```

## Checking influential points

```{r }

# Check for influential points using Cook's distance
cooks_d_trash_log <- cooks.distance(trash_model_log)

# Visualize Cook's Distance
cooks_data_trash_log <- data.frame(Observation = 1:length(cooks_d_trash_log), Cooks_Distance = cooks_d_trash_log)
ggplot(cooks_data_trash_log, aes(x = Observation, y = Cooks_Distance)) +
  geom_point() +
  geom_hline(yintercept = 4 / length(cooks_d_trash_log), color = "red", linetype = "dashed") +
  labs(title = "Cook's Distance for Influential Observations (Log Model)",
       x = "Observation",
       y = "Cook's Distance") +
  theme_minimal()

# Identify observations with Cook's distance greater than 4/n
influentials_trash_log <- which(cooks_d_trash_log > (4 / nrow(trashwheel)))
influentials_trash_log

```


## Checking Normality of Residuals
```{r }
# Calculate residuals
residuals_trash_log <- residuals(trash_model_log)

# Plot histogram of residuals
ggplot(data.frame(residuals_trash_log = residuals_trash_log), aes(x = residuals_trash_log)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "blue", alpha = 0.7) +
  labs(title = "Histogram of Residuals (Log Model)", x = "Residuals", y = "Frequency") +
  theme_minimal()

# Q-Q Plot for residuals
qqnorm(residuals_trash_log)
qqline(residuals_trash_log, col = "red", lwd = 2)

```

## Checking Linearity
```{r }
# Get the fitted values
fitted_values_trash_log <- fitted(trash_model_log)

# Create a data frame for residuals vs fitted values
residuals_vs_fitted_df_trash_log <- data.frame(
  fitted_values_trash_log, 
  residuals_trash_log = residuals(trash_model_log)
)

# Residuals vs Fitted plot
ggplot(residuals_vs_fitted_df_trash_log, aes(x = fitted_values_trash_log, y = residuals_trash_log)) +
  geom_point(color = "skyblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  labs(title = "Residuals vs Fitted Values (Log Model)", x = "Fitted Values", y = "Residuals") +
  theme_minimal()


```


## Checking homoscedasticity
```{r }
# Scale-Location Plot for Visual Check
plot(trash_model_log, which = 3)

# Breusch-Pagan Test for Homoscedasticity
bptest_trash_log <- bptest(trash_model_log)
bptest_trash_log

```

## Checking Multicolinearity
```{r }
# Calculate and check the VIF for the model
vif_values_trash_log <- vif(trash_model_log)

# Print VIF values
print(vif_values_trash_log)

```



## Building a more complex model where season is included
```{r }

# Create the Season variable based on Month
trashwheel$Season <- case_when(
  trashwheel$Month %in% c("december", "january", "february") ~ "Winter",
  trashwheel$Month %in% c("march", "april", "may") ~ "Spring",
  trashwheel$Month %in% c("june", "july", "august") ~ "Summer",
  trashwheel$Month %in% c("september", "october", "november") ~ "Autumn",
  TRUE ~ NA_character_
)

# Convert Season to a factor variable
trashwheel$Season <- factor(trashwheel$Season, levels = c("Winter", "Spring", "Summer", "Autumn"))

# Verify the new variable
table(trashwheel$Season)

# Ensure Month is lowercase to avoid mismatches
trashwheel$Month <- tolower(trashwheel$Month)

# Create the Season variable based on Month
trashwheel$Season <- case_when(
  trashwheel$Month %in% c("december", "january", "february") ~ "Winter",
  trashwheel$Month %in% c("march", "april", "may") ~ "Spring",
  trashwheel$Month %in% c("june", "july", "august") ~ "Summer",
  trashwheel$Month %in% c("september", "october", "november") ~ "Autumn",
  TRUE ~ NA_character_
)

# Convert Season to a factor variable
trashwheel$Season <- factor(trashwheel$Season, levels = c("Winter", "Spring", "Summer", "Autumn"))

# Verify the new variable
table(trashwheel$Season)

# Linear regression model: Predict Weight by also Season
season_model <- lm(Weight ~ PlasticBottles + Polystyrene + CigaretteButts + 
                  GlassBottles + PlasticBags + Wrappers + SportsBalls + Season, data = trashwheel)

# View the model summary
summary(season_model)


```


#### Checking influential points

```{r }

# Check for influential points using Cook's distance
cooks_d <- cooks.distance(season_model)

# Visualize Cook's Distance
cooks_data <- data.frame(Observation = 1:length(cooks_d), Cooks_Distance = cooks_d)
ggplot(cooks_data, aes(x = Observation, y = Cooks_Distance)) +
  geom_point() +
  geom_hline(yintercept = 4 / length(cooks_d), color = "red", linetype = "dashed") +
  labs(title = "Cook's Distance for Influential Observations",
       x = "Observation",
       y = "Cook's Distance") +
  theme_minimal()

# Identify observations with Cook's distance greater than 4/n
influentials <- which(cooks_d > (4 / nrow(trashwheel)))
influentials

```

#### Checking normality

```{r }
# Calculate residuals
residuals_model <- residuals(season_model)

# Plot histogram
ggplot(data.frame(residuals = residuals_model), aes(x = residuals)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "blue", alpha = 0.7) +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Frequency") +
  theme_minimal()


```

#### Checking linearity

```{r }

# Get the fitted values
fitted_values <- fitted(season_model)

# Create a data frame for ggplot
residuals_vs_fitted_df <- data.frame(fitted_values, residuals = residuals(season_model))

# Residuals vs Fitted plot
ggplot(residuals_vs_fitted_df, aes(x = fitted_values, y = residuals)) +
  geom_point(color = "skyblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  labs(title = "Residuals vs Fitted Values-ordinal", x = "Fitted Values", y = "Residuals") +
  theme_minimal()


```

#### Checking for homoscedasticity
```{r }
plot(season_model, which = 3)

# Breusch-Pagan Test for homoscedasticity
bptest(season_model)

```


#### Checking the multicollinearity assumption

```{r}

# Calculate and check the VIF for the complex model
vif_values <- vif(season_model)

# Print VIF values
print(vif_values)

```

#### Modifying the model
```{r }

# List of indices to remove
influential_indices <- c(13, 19, 20, 26, 30, 34, 50, 62, 67, 94, 107, 113, 117, 122, 130, 
                         272, 329, 407, 486, 518, 574, 578, 631, 633, 638, 639, 644, 645, 
                         647, 651, 657, 658, 662, 673, 696, 709, 716, 726, 749, 750, 754, 
                         755, 756, 770, 830)

# Remove these rows from the trashwheel dataset
trashwheel_cleaned <- trashwheel %>%
  filter(!row_number() %in% influential_indices)

# Check the first few rows to ensure the removal
head(trashwheel_cleaned)


# building the logarthimical mdoel
season_model_log <- lm(log_Weight ~ PlasticBottles + Polystyrene + CigaretteButts + GlassBottles + PlasticBags + Wrappers + SportsBalls + Season, data = trashwheel)

# summary of the model
summary(season_model_log)


```

#### Checking influential points

```{r }
# Check for influential points using Cook's distance
cooks_d_log <- cooks.distance(season_model_log)

# Visualize Cook's Distance
cooks_data_log <- data.frame(Observation = 1:length(cooks_d_log), Cooks_Distance = cooks_d_log)
ggplot(cooks_data_log, aes(x = Observation, y = Cooks_Distance)) +
  geom_point() +
  geom_hline(yintercept = 4 / length(cooks_d_log), color = "red", linetype = "dashed") +
  labs(title = "Cook's Distance for Influential Observations",
       x = "Observation",
       y = "Cook's Distance") +
  theme_minimal()

# Identify observations with Cook's distance greater than 4/n
influentials_log <- which(cooks_d_log > (4 / nrow(trashwheel_cleaned)))
influentials_log

```

#### Checking normality
```{r }
# Calculate residuals for the log model
residuals_model_log <- residuals(season_model_log)

# Plot histogram
ggplot(data.frame(residuals = residuals_model_log), aes(x = residuals)) +
  geom_histogram(binwidth = 0.5, color = "black", fill = "blue", alpha = 0.7) +
  labs(title = "Histogram of Residuals", x = "Residuals", y = "Frequency") +
  theme_minimal()

```

#### Checking linearity
```{r }
# Get the fitted values for the log model
fitted_values_log <- fitted(season_model_log)

# Create a data frame for ggplot
residuals_vs_fitted_df_log <- data.frame(fitted_values_log, residuals = residuals(season_model_log))

# Residuals vs Fitted plot
ggplot(residuals_vs_fitted_df_log, aes(x = fitted_values_log, y = residuals)) +
  geom_point(color = "skyblue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  labs(title = "Residuals vs Fitted Values-ordinal", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

```


#### Checking homoscedasticity
```{r }

plot(season_model_log, which = 3)

# Breusch-Pagan Test for homoscedasticity for the log model
bptest(season_model_log)

```


#### checking multicolinearity
```{r }

# Calculate and check the VIF for the log model
vif_values_log <- vif(season_model_log)

# Print VIF values
print(vif_values_log)

```


## Writing the final models
```{r }

# the modfied version of trash_model
final_trash <- lm(log_Weight ~ PlasticBottles + Polystyrene + CigaretteButts + 
                  GlassBottles + PlasticBags + Wrappers + SportsBalls, 
                  data = trashwheel) 

# Modified version of season_model
final_season<- lm(log_Weight ~ PlasticBottles + Polystyrene + CigaretteButts + 
                  GlassBottles + PlasticBags + Wrappers + SportsBalls + Season, 
                  data = trashwheel)

```


### Comparing the models

```{r }

# Compare AIC values of the two models
AIC(final_trash, final_season)
# Compare BIC values of the two models
BIC(final_trash, final_season)


summary(final_trash)

summary(final_season)

```

# Results and discussion
## The regression equation
for the simple model (final_trash) predicting the log of weight based on different trash categories is as follows:
log ⁡ ( Weight ) = 1.133 − 0.00001963 × PlasticBottles − 0.00002926 × Polystyrene + 0.00000309 × CigaretteButts − 0.001765 × GlassBottles + 0.00004400 × PlasticBags − 0.00004355 × Wrappers + 0.002846 × SportsBalls 

### Model Comparison
Two regression models were used to predict the log-transformed weight of trash: the simpler model (final_trash) and the more complex model (final_season) that includes an additional variable for seasons. Simple Model (final_trash): The simple model explained a small proportion of the variance in the log-transformed weight of trash, with an adjusted R-squared of 0.1506. The model had a significant F-statistic, 𝐹 ( 7 , 985 ) = 26.12 F(7,985)=26.12, 𝑝 < 2.2 𝑒 − 16 p<0.001, suggesting that the included variables were collectively significant predictors. Among the variables, PlasticBottles, Polystyrene, CigaretteButts, PlasticBags, Wrappers, and SportsBalls were significant predictors of log-transformed weight. GlassBottles, however, was not a significant predictor (p = 0.11661). Complex Model (final_season): The more complex model, which also includes the variable Season, accounted for a slightly larger proportion of variance in the log-transformed weight of trash, with an adjusted R-squared of 0.1564. The model's overall fit was significant, with 𝐹 ( 10 , 982 ) = 19.39 F(10,982)=19.39, 𝑝 < 0,001. In this model, PlasticBottles, Polystyrene, CigaretteButts, PlasticBags, Wrappers, and SportsBalls remained significant, similar to the simple model. Additionally, SeasonSpring and SeasonSummer were found to be significant predictors, with SeasonAutumn showing no significant effect (p = 0.216571). 

### Model Fit and Complexity 

Although the complex model explains slightly more variance in the dependent variable (log-transformed weight of trash), with an adjusted R-squared increase from 0.1506 to 0.1564, the improvement is relatively small. This suggests that while seasonality may have a small effect on the weight of trash, the inclusion of season variables doesn't significantly improve the model compared to the simpler one.

## Conclusion

The simple model (final_trash) is preferred due to the simplicity and the fact that the addition of seasonal information in the complex model (final_season) does not substantially increase the explanatory power of the model. Removing the seasonality factor does not negatively impact the model's predictive accuracy much, making the simpler model an effective choice in this case.